---
title: "Analysis on CDK4/6i + BETi +  Combo Synergy"
output: html_document
date: "2025-11-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Libraries 
```{r}
suppressPackageStartupMessages({
  library(ggplot2);library(ggrepel);library(ggpubr);
  library(data.table);library(RColorBrewer);library(tidyverse);
  library(preprocessCore);library(future.apply);library(DESeq2);
  library(pheatmap);library(sva);library(viridis);
  library(limma);library(emmeans);library(broom);
  library(janitor);library(tidyplots);library(dplyr);
  library(writexl);library(scToppR);library(fgsea);
  library(clusterProfiler);library(org.Hs.eg.db);library(dplyr);library(readxl)
})
```

##############
#------------- Loading counts and processing
```{r}
gene_counts<- read_xlsx("edgeR_new_exploration/counts.xlsx")
counts<- gene_counts |> filter(gene_biotype == "protein_coding") |>
  dplyr::select(gene_name, 2:13) |>
  dplyr::select(
    gene_name,
    DMSO_1, DMSO_2, DMSO_3,
    BMS_2, BMS_3, BMS_4,
    Ribo_1, Ribo_2, Ribo_3,
    Combo_1, Combo_3, Combo_4
  )
rownames(counts)<- NULL

p_exp <- counts |>
  dplyr::distinct(gene_name, .keep_all = TRUE)
rownames(p_exp)<- NULL
p_exp <- p_exp |> column_to_rownames('gene_name')

# Reading metadata
metadata<- read.csv("edgeR_new_exploration/metadata.csv")
metadata<- metadata |> column_to_rownames("X")
metadata$Condition<-factor(trimws(metadata$Condition))
metadata$Condition<-factor(metadata$Condition, levels = c("DMSO","BMS", "Ribo", "Combo"))
levels(metadata$Condition)

all(colnames(p_exp) == rownames(metadata))
```
##############
#------------- CPM normalization
```{r}
## Converting to cpms: Counts per million 
cpm <-future_apply(p_exp, 2, function(x) x/sum(as.numeric(x)) * 10^6) 

## Filter low cpm counts for both conditions 
filter<-apply(cpm, 1, function(x) all(x[1:3]>=0.5) | all(x[4:6]>=0.5) | all(x[7:9]>=0.5) | all(x[10:12]>=0.5))
counts_filt <- p_exp[filter,]  ## 13772 genes
cpm_filt <- cpm[filter,]   ## 13772 genes

#### Saving cpm and counts filter
cpm_filt_xlsx<-cpm_filt %>% as.data.frame() %>% tibble::rownames_to_column(var = "Gene")
write_xlsx(cpm_filt_xlsx, "edgeR_new_exploration/deg/gene_cpm.xlsx")

counts_filt_xlsx<-counts_filt %>%as.data.frame() %>%tibble::rownames_to_column(var = "Gene")
write_xlsx(counts_filt_xlsx, "edgeR_new_exploration/deg/gene_filt_counts.xlsx")

## Applying log_transformation
logCPM<- log2(cpm_filt +1)
boxplot(logCPM)

## Normalizing quantiles from cpm counts
p<- normalize.quantiles(as.matrix(logCPM))
rownames(p)<-rownames(logCPM)
colnames(p) <-colnames(logCPM)
boxplot(p)

## Performing PCA using the quantile normalization from the CPMs and Visualize
metadata$Subject_ID<- rownames(metadata)
pca.Samples<-prcomp(t(p))
PCi<-data.frame(pca.Samples$x, Condition= metadata$Condition, ID= metadata$Subject_ID )
eig <- (pca.Samples$sdev)^2 ### Calculating the eigenvalue
variance <- eig*100/sum(eig) ## Calculating variance
pdf("edgeR_new_exploration/deg/PCA_cpms.pdf",width=7,height=5,useDingbats=FALSE)
ggscatter(
  PCi,
  x = "PC1",
  y = "PC2",
  color = "Condition",
  palette = c("#D8CE7A", "#A5D996", "#8259D9", "blue4"),
  size = 6,
  ellipse = TRUE,
  ellipse.type = "confidence",
  ellipse.alpha = 0.2
) +
  xlab(paste("PC1 (", round(variance[1],1), "% )")) +
  ylab(paste("PC2 (", round(variance[2],1), "% )")) +
  labs(color = "Condition", fill = "Condition") +
  scale_color_manual(
    values = c("#D8CE7A", "#A5D996", "#8259D9", "blue4"),
    labels = c("DMSO", "BETi", "CDK4/6i", "Combo")
  ) +
  scale_fill_manual(
    values = c("#D8CE7A", "#A5D996", "#8259D9", "blue4"),
    labels = c("DMSO", "BETi", "CDK4/6i", "Combo")
  ) +
  theme_classic() +
  theme(
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 16),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12.5),
    axis.ticks = element_line(linewidth = 0.8),
    axis.ticks.length = unit(0.2, "cm"))
dev.off()
```
##############
#------------- Differential expression analysis using DESEQ2
```{r}
### Setting metadata conodition as : DMSO as reference
metadata$Condition<- factor(metadata$Condition)
metadata$Condition <- relevel(metadata$Condition, ref = "DMSO")

## Creating DESEq object 
dds<- DESeqDataSetFromMatrix(countData = p_exp,colData = metadata, design = ~ Condition)
#counts(dds)
## Estimating factors
dds<-estimateSizeFactors(dds)
## Performing DESEq
dds<-DESeq(dds, full = design(dds),betaPrior = FALSE)
resultsNames(dds)

### #####
####--------- Obtaining resulst 
cd <- "Condition_Combo_vs_DMSO"  #--< Changing according to purposes
res = results(dds, name = cd)
head(res[order(res$padj),]) ## order to FDR
#### Saving
CTX_FullTab <-as.data.frame(results(dds, name = cd, cooksCutoff = F, independentFiltering = F))%>% rownames_to_column("Gene")
CTX_DGE <- CTX_FullTab %>%
  mutate(Abs = abs(log2FoldChange)) %>%
  filter(padj < 0.05 & Abs > 0.3) %>% 
  arrange(desc(Abs))

### Saving 
## Establishing filepath --< Changing accoridng to purposes
file_path<- "edgeR_new_exploration/Refinment/deg/Combo_BMS/"

write_xlsx(CTX_FullTab, file.path(file_path, "CTX_FullTab.xlsx"))
write_xlsx(CTX_DGE, file.path(file_path, "CTX_DGE.xlsx"))

###################
### Shrinkage
res_shrink<-lfcShrink(dds, coef = cd, type = "ashr" )
degs_shrink<- subset(res_shrink, padj < 0.05 & abs(log2FoldChange) > 0.3 )

### Getting dataframes
CTX_FullTab_shrink<- as.data.frame(res_shrink) %>%rownames_to_column("Gene")
CTX_DGE_shrink<-as.data.frame(degs_shrink) %>%rownames_to_column("Gene")

write_xlsx(CTX_FullTab_shrink, file.path(file_path, "CTX_FullTab_shrink.xlsx"))
write_xlsx(CTX_DGE_shrink, file.path(file_path, "CTX_DGE_shrink.xlsx"))

save(CTX_FullTab, CTX_DGE, CTX_FullTab_shrink, CTX_DGE_shrink, metadata, metadata, 
     file = file.path(file_path, "CTX_Dge_Data.RData"))

save(dds, file = file.path(file_path, "Ref_DMSO.RData"))
```


######### Loading one by one exploring each dataset 
###### Visualizations of the shrinkage data
##### Volcano plots ##### ##### #####
```{r}
### Loading one by one
data<- read_xlsx("edgeR_new_exploration/deg/Combo_Ribo/CTX_FullTab_shrink.xlsx")
## Creatimng temporary dataframe
df <- unique_BMS %>%
  mutate(LOG = -log10(padj), ABS = abs(log2FoldChange)) %>%
  mutate(Threshold = if_else(padj < 0.05 & ABS > 0.3, "TRUE","FALSE")) %>%
  mutate(Direction = case_when(log2FoldChange > 0.3 & padj < 0.05 ~ "UpReg", log2FoldChange < -0.3 & padj < 0.05 ~ "DownReg")) ### why 0.3
dim(df)

write_xlsx(unique_Ribo, "edgeR_new_exploration/deg/Unique_Ribo/unique_Ribo_DEGs.xlsx")
#df<- df |> filter(padj > 0)
### Top genes based on Direction and padj
### These genes are the most statistically significance
top_labelled <- df %>%
  group_by(Direction) %>%
  na.omit() %>%
  arrange(padj) %>%
  top_n(n = 7, wt = LOG)

head(top_labelled)

pdf("edgeR_new_exploration/deg/Unique_Ribo/Volcano_Plot_CTX.pdf",width=6,height=6,useDingbats=FALSE)
ggscatter(df,
          x="log2FoldChange",
          y="LOG",
          color = "Direction",
          palette=c("dodgerblue1","firebrick1"),
          size = 1,
          alpha=0.3,
          shape=19)+
  xlab("log2(Fold Change)")+
  ylab("-log10(FDR)")+
  geom_vline(xintercept = 0, colour = "grey",linetype="dotted",size=1,alpha=0.5) +
  geom_vline(xintercept = 0.3, colour = "black",linetype="dotted",size=1,alpha=0.5) +
  geom_vline(xintercept = -0.3, colour = "black",linetype="dotted",size=1,alpha=0.5) +
  geom_hline(yintercept = 1.3, colour = "grey",linetype="dotted",size=1,alpha=0.5) +
  geom_text_repel(data = top_labelled,
                  mapping = aes(label = Gene),
                  size = 4,
                  box.padding = unit(0.4, "lines"),
                  point.padding = unit(0.4, "lines"))+
  theme(legend.position="none")+
  ylim(0,50) + xlim(-8,+8) 
dev.off()


#### Gene ontology
### prioritize LOgfc ::100 up and down 
```
### Gene ontology #######
################ ########
```{r}
library(scToppR)
tmp1<- df|> filter(Direction %in% c("UpReg", "DownReg"))
toppData <-  toppFun(
       tmp1,
       topp_categories = NULL,
       cluster_col = "Direction",
       gene_col = "Gene",
       p_val_col = "padj",
       logFC_col = "log2FoldChange",
       pval_cutoff = 0.05,
       min_genes = 10,
       max_genes = 500,
       max_results = 50
     )
toppPlot(
       toppData,
       category = "GeneOntologyMolecularFunction",
       num_terms = 10,
       p_val_adj = "BH",
       p_val_display = "log",
       save = TRUE,
       #save_dir = "edgeR_new_exploration/deg/Unique_Ribo/GO/pseudobulk_moderate_GO",
       width = 5,
       height = 6
     )
```
```{r}
toppPlot(
       toppData,
       category = "GeneOntologyBiologicalProcess",
       num_terms = 10,
       p_val_adj = "BH",
       p_val_display = "log",
       save = TRUE,
       #save_dir = "edgeR_new_exploration/deg/unique_Ribo/GO/pseudobulk_moderate_GO",
       width = 5,
       height = 6
     )
```
```{r}
save(toppData, file = "edgeR_new_exploration/deg/Unique_Ribo/GO/pseudobulk_moderate_GO/TopData.RData")
GO<- toppData |> as.data.frame()
library(writexl)
write_xlsx(GO, "edgeR_new_exploration/deg/Unique_Ribo/GO/pseudobulk_moderate_GO/GO.xlsx")
#write_xlsx(unique_Combo, "edgeR_new_exploration/deg/Unique_Ribo/Combo_unique_CTX_DGE_shrink.xlsx")
```


#### Viz for AMPK- signaling
##### Volcano plot is not working ####
###### ############### ###############
```{r}
library(readxl)
Combo_DMSO<- read_xlsx("edgeR_new_exploration/deg/Combo_DMSO/CTX_FullTab_shrink.xlsx")
## Volcano plot 
 
top_labelled <- df %>%group_by(Direction) %>% filter(!Direction == "NS") %>% 
  arrange(padj) %>%top_n(n = 10, wt = LOG)

amp_genes<- c("Aak1", "Araf", " Bcr", " Cdc42bpa", " Mtor", "Dclk1",
              "Ern1", "Gsk3b", "Lmtk2", "Mtor", "Pikfyve", "Prkaa1",
              "Prkag1", "Prkx", "Pskh1", "Ripk2", "Rps6ka2", "Sik2",
              "Srpk2", "Stk36", "Taok2", "Trio", "Ttbk1", 'Ttbk2',
              "Wnk3", 'Wnk4'
              )

top_labelled <- df[df$Gene %in% amp_genes,]

ggplot(df, aes(x = log2FoldChange, y = LOG)) +
  theme_classic(base_size = 13) +
  theme(
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    axis.text.x  = element_text(size = 16),
    axis.text.y  = element_text(size = 16),
    axis.ticks   = element_line(linewidth = 0.8),
    axis.ticks.length = unit(0.25, "cm")
  ) +
  geom_point(aes(color = Direction), size = 1, alpha = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c(DownReg = "#5CACDB", NS = "grey", UpReg = "#EA7FA3")) +

  ggnewscale::new_scale_color() +

  geom_text_repel(
    data = top_labelled,
    aes(label = Gene, color = Direction),
    max.overlaps = Inf,
    size = 5,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c(DownReg = "#2C78A6", UpReg = "#C64A6E")) +

  geom_vline(xintercept = 0, linetype = 2, color = "red") +
  geom_hline(yintercept = -log10(0.05), linetype = 2, color = "red") +
  xlab("avg_log2FC") +
  ylab("-log10(adj p-value)") +
  scale_x_continuous(limits = c(-10, 10))

### Topdata just one 
load("edgeR_new_exploration/deg/Unique_combo/GO/pseudobulk_moderate_GO/TopData.RData")
x<- toppData
x<- x |> filter(Name %in% c("AMP-activated protein kinase activity",
                            "insulin receptor binding"
                            ))
toppPlot(
       x,
       category = "GeneOntologyMolecularFunction",
       num_terms = 4,
       p_val_adj = "BH",
       p_val_display = "log",
       save = TRUE,
       save_dir = "edgeR_new_exploration/deg/Unique_combo/AMPK_signaling/",
       width = 6,
       height = 3.5
     )

```



############
### Heatmaps #####------
############
```{r}
### Laoding the normalized matrix and the DEGs
#CTX_DGE_shrink<- openxlsx::read.xlsx("deg/CTX_DGE_shrink.xlsx")
counts_filt_xlsx<- openxlsx::read.xlsx("edgeR_new_exploration/deg/gene_cpm.xlsx")
#metadata<- read.table("deg/metadata.txt")

library(viridis)
p_exp<- counts_filt_xlsx
p_exp<- p_exp |> column_to_rownames("Gene")
mat <- p_exp[rownames(p_exp)%in% amp_genes,] 

dim(mat)

anno<-metadata
anno$Subject_ID<- NULL
anno$Condition <- factor(anno$Condition, levels = c("DMSO", "Ribo", "BMS", "Combo"))
my_colour = list(
    Condition = c(DMSO = "#4CAF50", Ribo = "#E91E63",
                  BMS = "darkblue", Combo = "darkred")
)
pdf("edgeR_new_exploration/deg/Combo_BMS/DEGs_Heatmap_CTX.pdf",width=6,height=7,useDingbats=FALSE)
pheatmap(mat,
         scale="row",
         show_rownames = F,
         annotation= anno,
         annotation_colors = my_colour,
         colorRampPalette(rev(viridis(100)))(100),
         cellwidth = 12,
         #cutree_cols = 2,
         angle_col = 45,
         cluster_cols = F
         )
dev.off()

### Creating a heatmap of the most uoregulated and downregylated genes
top_up <- unique_Combo[order(unique_Combo$log2FoldChange, decreasing = TRUE), ][1:20, ]
top_down <- unique_Combo[order(unique_Combo$log2FoldChange, decreasing = FALSE), ][1:20, ]
top_bind <- rbind(top_up, top_down)

### MODIFYING THE P_exp to have just the Combo vs DMSO
q<- p_exp |> dplyr::select(1:3, 10:12)
mat <- q[rownames(q)%in% top_bind$Gene,]
anno<-metadata
anno$Subject_ID<- NULL
anno$Condition <- factor(anno$Condition, levels = c("DMSO", "Ribo", "BMS", "Combo"))
my_colour = list(
    Condition = c(DMSO = "#4CAF50", Ribo = "#E91E63",
                  BMS = "darkblue", Combo = "darkred")
)
anno$Sex<- NULL
pdf("edgeR_new_exploration/deg/Synergy_modeling/Top0.1Heatmap_CTX.pdf",width=7,height=4,useDingbats=FALSE)
pheatmap(mat,
         scale="row",
         show_rownames = TRUE,
         annotation= anno,
         annotation_colors = my_colour,
         cutree_cols = 2,
         cellwidth = 20,
         cellheight = 8.5,
         border_color = "black",
         angle_col = 45,
         cluster_cols = F
         )
dev.off()


### Heatmap pf the AMPK signaling 
mat <- p_exp[rownames(p_exp)%in% amp_genes,]
q<- mat |> dplyr::select(1:3, 10:12)
q<- as.matrix(q)

metadata<- read_csv("edgeR_new_exploration/metadata.csv")
anno<-metadata
anno<- anno |> column_to_rownames('...1')
anno$Sex <- NULL
#anno$Subject_ID<- NULL
anno$Condition <- factor(anno$Condition, levels = c("DMSO", "Ribo", "BMS", "Combo"))
#my_colour = list(
#    Condition = c(DMSO = "#4CAF50", Ribo = "#E91E63",
#                  BMS = "darkblue", Combo = "darkred")
#)

my_colour = list(
    Condition = c(DMSO = "#4CAF50", Combo = "darkred")
)
pdf("edgeR_new_exploration/deg/Unique_combo/AMP_KHeatmap_CTX.pdf",width=7,height=5,useDingbats=FALSE)
pheatmap(q,
         scale="row",
         show_rownames = TRUE,
         annotation= anno,
         annotation_colors = my_colour,
         cutree_cols = 2,
         cellwidth = 20,
         cellheight = 8.5,
         border_color = "black",
         angle_col = 45,
         cluster_cols = T
         )
dev.off()
```

#### GSEA ####
##############
```{r}
run_gsea_per_cluster <- function(df,
                                 pathway_list,
                                 gene_col    = "feature",
                                 stat_col    = "logFC",
                                 cluster_col = "group",
                                 output_dir  = "deg/GSEA/",
                                 nperm       = 10000,
                                 minSize     = 10,
                                 maxSize     = 500,
                                 padj_cutoff = 0.05,
                                 label_prefix = "contrast") {

  dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

  clusts <- sort(unique(df[[cluster_col]]))
  message("Running GSEA for ", length(clusts), " clusters: ",
          paste(clusts, collapse = ", "))

  per_cluster <- map(clusts, function(cl) {
    sub <- df %>%
      filter(.data[[cluster_col]] == cl) %>%
      filter(!is.na(.data[[stat_col]])) %>%
      # If a gene appears multiple times within a cluster, keep the largest |stat|
      arrange(desc(abs(.data[[stat_col]]))) %>%
      distinct(.data[[gene_col]], .keep_all = TRUE) %>%
      arrange(desc(.data[[stat_col]]))

    stat_vec <- sub[[stat_col]]
    names(stat_vec) <- sub[[gene_col]]

    if (length(stat_vec) < minSize) {
      warning("Cluster ", cl, ": fewer than minSize genes (", length(stat_vec), "). Skipping.")
      return(tibble())
    }

    set.seed(0708)
    fg <- fgsea(pathways = pathway_list,
                stats    = stat_vec,
                nperm    = nperm,
                minSize  = minSize,
                maxSize  = maxSize)

    res <- as_tibble(fg) %>%
      arrange(padj, desc(NES)) %>%
      mutate(cluster = cl, label = label_prefix)

    # Save per-cluster results
    write_csv(res, file.path(output_dir, sprintf("%s_cluster-%s_GSEA.csv", label_prefix, cl)))

    # Also save significant up/down
    up   <- res %>% filter(padj < padj_cutoff, ES > 0)
    down <- res %>% filter(padj < padj_cutoff, ES < 0)
    write_csv(up,   file.path(output_dir, sprintf("%s_cluster-%s_GSEA_up.csv",   label_prefix, cl)))
    write_csv(down, file.path(output_dir, sprintf("%s_cluster-%s_GSEA_down.csv", label_prefix, cl)))

    res
  })

  combined <- bind_rows(per_cluster)

  # Save combined table (all clusters)
  write_csv(combined, file.path(output_dir, sprintf("%s_ALL_clusters_GSEA.csv", label_prefix)))

  invisible(list(per_cluster = per_cluster, combined = combined))
}
#BiocManager::install("fgsea")
library(fgsea)
hallmark_pathways <- gmtPathways("edgeR_new_exploration/deg/BMS_DMSO/GSEA/mh.all.v2024.1.Mm.symbols.gmt")

df$cluster<- "CombovsRibo"
res <- run_gsea_per_cluster(
  df           = df,
  pathway_list = hallmark_pathways,
  gene_col     = "Gene",
  stat_col     = "log2FoldChange",
  cluster_col  = "cluster",
  output_dir   = "edgeR_new_exploration/deg/Combo_Ribo/GSEA/",
  nperm        = 10000,
  label_prefix = "Combo_vs_Ribo"
)

res$combined %>% 
  arrange(padj) %>% 
  head(50)

### Saving GSEA output 
gsea_output1<-res$combined |>
  as.data.frame()
gsea_output1$leadingEdge <- sapply(gsea_output1$leadingEdge, function(x) paste(x, collapse = ", "))
write.csv(gsea_output1, "edgeR_new_exploration/deg/Combo_Ribo/GSEA/full_genes_list_GSEA.csv")
save(res, file = 'edgeR_new_exploration/deg/Combo_Ribo/GSEA/GSEA.RData')
```

```{r}
plot_bulk_gsea_barplot_by_cluster <- function(gsea_df,
                                              cluster_col = "group",
                                              output_dir = "deg/GSEA/",
                                              label = "sCRNAseq") {
  library(ggpubr)
  library(ggplot2)

  # Create output dir
  dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

  clusts <- sort(unique(gsea_df[[cluster_col]]))

  for (cl in clusts) {
    # Subset this cluster
    df_cl <- gsea_df[gsea_df[[cluster_col]] == cl, , drop = FALSE]

    # Preprocess for plotting (same as your function)
    tpm <- df_cl %>%
      mutate(adjPvalue = ifelse(padj <= 0.05, "significant", "non-significant"),
             pathway = gsub("HALLMARK_", "", pathway))  # Clean pathway names if Hallmark

    # Define colors (same)
    cols <- c("non-significant" = "black", "significant" = "#CD4071FF")

    # Create barplot for downregulated pathways (ES < 0) — same format
    p <- ggbarplot(tpm %>% filter(ES > 0),
                   x = "pathway", y = "NES",
                   fill = "adjPvalue",
                   color = "white",
                   palette = cols,
                   sort.val = "asc",
                   sort.by.groups = FALSE,
                   x.text.angle = 90,
                   ylab = "NES",
                   legend.title = "FDR < 0.05",
                   rotate = TRUE,
                   ggtheme = theme_minimal()) +
      theme(
        axis.text.x = element_text(size = 18, color = "black"),
        axis.title.x = element_text(size = 16, color = "black", face = "bold"),
        axis.text.y = element_text(size = 14, color = "black"),
        axis.title.y = element_text(size = 16, color = "black", face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14)
      )

    # Save plot (one file per cluster)
    ggsave(
      filename = file.path(output_dir, paste0(label, "_cluster-", cl, "_Upregulated_GSEA.jpeg")),
      plot = p,
      width = 9,
      height = 9,
      dpi = 600,
      units = "in"
    )
  }
}

plot_bulk_gsea_barplot_by_cluster(
  gsea_df   = res$combined,   # or your per-cluster GSEA tibble
  cluster_col = "cluster",    # use "ClusterID" if that's your column
  output_dir = "edgeR_new_exploration/deg/Combo_Ribo/GSEA/barplots",
  label      = "Combo_Ribo"
)
```

### Obtaing synergy 
```{r}
p_exp
metadata1
metadata2 <- metadata %>%
  mutate(
    BMS_treat  = ifelse(Condition %in% c("BMS", "Combo"), 1, 0),
    Ribo_treat = ifelse(Condition %in% c("Ribo", "Combo"), 1, 0)
  )
metadata2$BMS_treat  <- factor(metadata2$BMS_treat)
metadata2$Ribo_treat <- factor(metadata2$Ribo_treat)
str(metadata2)
table(metadata2$Condition, metadata2$BMS_treat, metadata2$Ribo_treat)

### Deseq
## Creating DESEq object 
dds2<- DESeqDataSetFromMatrix(countData = p_exp,
                             colData = metadata2,
                             design = ~ BMS_treat * Ribo_treat) # ~ BMS_treat * Ribo_treat 
counts(dds2)

## Estimating factors
dds2<-estimateSizeFactors(dds2)
## Performing DESEq
dds2<-DESeq(dds2, 
           full = design(dds2),
           betaPrior = FALSE)
resultsNames(dds2)

#res<-results(dds, contrast = c("Condition","DMSO","BMS", "Ribo", "Combo"))
### Exploration ###### ###########################
res2 = results(dds2, name = "BMS_treat1.Ribo_treat1")
head(res2[order(res2$padj),]) ## order to FDR
### Subsetting the number of DEGs based on 1
#deg<-subset(res, padj < 0.05 & abs(log2FoldChange) > 1 )


#### saving this as excel file 
CTX_FullTab2 <-as.data.frame(results(dds2, name = "BMS_treat1.Ribo_treat1", cooksCutoff = F, independentFiltering = F))%>%
  rownames_to_column("Gene")
CTX_DGE2 <- CTX_FullTab2 %>%
  mutate(Abs = abs(log2FoldChange)) %>%
  filter(padj < 0.1 & Abs > 0) %>%   ### FDR<0.05 and absolute value of 3 ( 100% up or down)
  arrange(desc(Abs))


CTX_DGE2$Gene
### Saving this 
library(writexl)
write_xlsx(CTX_FullTab2, "edgeR_new_exploration/deg/Synergy_modeling/CTX_FullTab.xlsx")
write_xlsx(CTX_DGE2, "edgeR_new_exploration/deg/Synergy_modeling/CTX_DGE.xlsx")
save(dds2, file = "edgeR_new_exploration/deg/Synergy_modeling/dds2.RData")
```

### Loading DEGs and investigating the effect of the combo ######
########################## ####################### ###############
```{r}
Ribo<- read_xlsx("edgeR_new_exploration/deg/Ribo_DMSO/CTX_DGE_shrink.xlsx")
BMS<- read_xlsx("edgeR_new_exploration/deg/BMS_DMSO/CTX_DGE_shrink.xlsx")
Combo<- read_xlsx("edgeR_new_exploration/deg/Combo_DMSO/CTX_DGE_shrink.xlsx")
Combo_BMS<- read_xlsx("edgeR_new_exploration/deg/Combo_BMS/CTX_DGE_shrink.xlsx")
Combo_Ribo<- read_xlsx("edgeR_new_exploration/deg/Combo_Ribo/CTX_DGE_shrink.xlsx")

RiboDEGs <- Ribo %>%
  mutate(LOG = -log10(padj), ABS = abs(log2FoldChange)) %>%
  mutate(Threshold = if_else(padj < 0.05 & ABS > 0.3, "TRUE","FALSE")) %>%
  mutate(Direction = case_when(log2FoldChange > 0.3 & padj < 0.05 ~ "UpReg", log2FoldChange < -0.3 & padj < 0.05 ~ "DownReg")) ### why 0.3
BMSDEGs <- BMS %>%
  mutate(LOG = -log10(padj), ABS = abs(log2FoldChange)) %>%
  mutate(Threshold = if_else(padj < 0.05 & ABS > 0.3, "TRUE","FALSE")) %>%
  mutate(Direction = case_when(log2FoldChange > 0.3 & padj < 0.05 ~ "UpReg", log2FoldChange < -0.3 & padj < 0.05 ~ "DownReg")) ### why 0.3
ComboDEGs <- Combo %>%
  mutate(LOG = -log10(padj), ABS = abs(log2FoldChange)) %>%
  mutate(Threshold = if_else(padj < 0.05 & ABS > 0.3, "TRUE","FALSE")) %>%
  mutate(Direction = case_when(log2FoldChange > 0.3 & padj < 0.05 ~ "UpReg", log2FoldChange < -0.3 & padj < 0.05 ~ "DownReg")) ### why 0.3
Combo_BMSDEGs <- Combo_BMS %>%
  mutate(LOG = -log10(padj), ABS = abs(log2FoldChange)) %>%
  mutate(Threshold = if_else(padj < 0.05 & ABS > 0.3, "TRUE","FALSE")) %>%
  mutate(Direction = case_when(log2FoldChange > 0.3 & padj < 0.05 ~ "UpReg", log2FoldChange < -0.3 & padj < 0.05 ~ "DownReg")) ### why 0.3
Combo_RiboDEGs <- Combo_Ribo %>%
  mutate(LOG = -log10(padj), ABS = abs(log2FoldChange)) %>%
  mutate(Threshold = if_else(padj < 0.05 & ABS > 0.3, "TRUE","FALSE")) %>%
  mutate(Direction = case_when(log2FoldChange > 0.3 & padj < 0.05 ~ "UpReg", log2FoldChange < -0.3 & padj < 0.05 ~ "DownReg")) ### why 0.3

### BMS driven
## Genes significantly changed in both BMS and Combo vs DMSO, not changed by Ribo, and Combo is not significantly different from BMS. → BMS is doing the heavy lifting.
BMS_driven <- intersect(BMSDEGs$Gene, ComboDEGs$Gene) # Intersect BMS and Combo
BMS_driven <- setdiff(BMS_driven, RiboDEGs$Gene) # exclude Ribo-driven
BMS_driven <- setdiff(BMS_driven, Combo_BMS$Gene) # Combo not different from BMS
BMS_driven <- setdiff(BMS_driven, CTX_DGE2$Gene) # keep synergy separate

### Ribo driven 
## Genes that change with Ribo and Combo, independent of BMS.
Ribo_driven <- intersect(RiboDEGs$Gene, ComboDEGs$Gene) # Intersect Ribo and Combo
Ribo_driven <- setdiff(Ribo_driven, BMSDEGs$Gene) # exclude BMS-driven
Ribo_driven <- setdiff(Ribo_driven, Combo_RiboDEGs$Gene) # Combo not different from Ribo
Ribo_driven <- setdiff(Ribo_driven, CTX_DGE2$Gene) # keep synergy separate

### Combo driven
## Changed only in Combo vs DMSO, not in either single-drug (within thresholds)
### Combo-specific stress responses
Combo_driven <- setdiff(ComboDEGs$Gene, union(BMSDEGs$Gene, RiboDEGs$Gene))
Combo_driven <- setdiff(Combo_driven, CTX_DGE2$Gene)  # synergy kept separate


### Venn diagram vs DMSO
library(ggVennDiagram)
gene_lists <- list(
  BMS   = BMSDEGs$Gene,
  Ribo  = RiboDEGs$Gene,
  Combo = ComboDEGs$Gene
)

ggVennDiagram(gene_lists)
library(ggvenn)
p4<-ggvenn(
  gene_lists, 
  fill_color = c("#0073C2FF", "#EFC000FF", "#CD534CFF"),
  stroke_size = 0.8, set_name_size = 4, show_percentage = F, text_size = 6,
  auto_scale = F
  )
p4

### Venn diagram vs driven genes 
library(ggVennDiagram)
gene_lists <- list(
  BMS_driven   = BMS_driven,
  Ribo_driven  = Ribo_driven,
  Combo_driven = Combo_driven
)
library(ggvenn)
p5<-ggvenn(
  gene_lists, 
  fill_color = c("#0073C2FF", "#EFC000FF", "#CD534CFF"),
  stroke_size = 0.8, set_name_size = 4, show_percentage = F, text_size = 6,
  auto_scale = F
  )
p5
```
##### Combo unique genes #####
############ #################
```{r}
all(Combo_driven %in% Ribo$Gene)
all(Combo_driven %in% BMS$Gene)
all(Combo_driven %in% Combo$Gene)

### Extracting unique Combo genes and BMS and RIbo 
unique_Combo<- Combo[Combo$Gene %in% Combo_driven,]
unique_BMS<- BMS[BMS$Gene %in% BMS_driven,]
unique_Ribo<- Ribo[Ribo$Gene %in% Ribo_driven,]
#### SAving them 

#### Gene ontology above 

#### Extracting the AMPK pathway genes 
a<-toppData[toppData$Name == "AMP-activated protein kinase activity",]
b <-a|> pull(Genes)
AMP_genes<- c(
  "Ern1","Rps6ka2","Cdc42bpa","Aak1","Lmtk2","Araf","Mtor","Stk36","Ripk2",
  "Srpk2","Bcr","Ttbk2","Ttbk1","Sik2","Gsk3b","Dclk1","Trio","Taok2",
  "Prkaa1","Prkag1","Prkx","Pskh1","Wnk4","Wnk3","Pikfyve")
x<- p_exp[rownames(p_exp) %in% AMP_genes,]
x <- x |> dplyr::select(1:3, 10:12)
str(x)


### Combo genes 
AMPK_genes<-Combo_unique[Combo_unique$Gene %in% AMP_genes,]
write_xlsx(AMPK_genes, "edgeR_new_exploration/deg/Unique_combo/AMP_uniq_combo.xlsx")

x1 <- x |>
  rownames_to_column("Gene") 

df_long <- x1 |>
  pivot_longer(
    cols = -Gene,
    names_to = "Sample",
    values_to = "Count"
  ) |>
  mutate(
    Condition = if_else(str_starts(Sample, "DMSO"), "DMSO", "Combo")
  )

# Summarise mean ± sd per gene & condition
df_summary <- df_long |>
  group_by(Gene, Condition) |>
  summarise(
    meanCount = mean(Count),
    sdCount   = sd(Count),
    .groups   = "drop"
  )
pdf("edgeR_new_exploration/deg/Unique_combo/AMP_signaling.pdf", width = 7, height = 5)
ggplot(df_summary, aes(x = Condition, y = meanCount, fill = Condition)) +
  geom_col(position = "dodge") +
  geom_errorbar(
    aes(ymin = meanCount - sdCount, ymax = meanCount + sdCount),
    width = 0.2
  ) +
  facet_wrap(~ Gene, scales = "free_y") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 8)
  ) +
  ylab("Mean log(CPMs +1)") +
  xlab("") +
  ggtitle("AMPK signalling genes")
dev.off()


df_long <- x |>
  rownames_to_column("Gene") |>
  pivot_longer(
    cols = -Gene,
    names_to = "Sample",
    values_to = "Expression"
  ) |>
  mutate(
    Condition = ifelse(grepl("^DMSO", Sample), "DMSO", "Combo")
  )

ggplot(df_long, aes(x = Condition, y = Expression)) +
  geom_violin(trim = FALSE) +
  geom_jitter(width = 0.1, size = 1, alpha = 0.7) +
  facet_wrap(~ Gene, scales = "free_y") +
  theme_bw() +
  labs(x = "", y = "Expression (normalized counts)")


write.csv(x1, file = "edgeR_new_exploration/deg/Unique_combo/AMP_signaling_cpms.csv")


### Creating a logFC bettwen plot of BMS vs Combo
BMSDEGs
ComboDEGs

# Join by gene
fc_scatter <- BMSDEGs %>%
  inner_join(ComboDEGs, by = "Gene")
```

##### Creating a cell signature ###########
############## ############ ###############

```{r}
library(Seurat)
library(AUCell)
load("C:/Users/amade/Documents/Ocasio_data/output/SeuObject_noDoub_integrated.RData")
DimPlot(seuObject_slim_nodoub, group.by = "celltype1")

### AUCell signature of Combo genes that are upregulated and downregulated 
Combo_up<- unique_Combo |> filter(log2FoldChange > 0) |> pull (Gene)
Combo_down<- unique_Combo |> filter(log2FoldChange < 0) |> pull (Gene)
### Creating signatures with AUCell
AUCell_list <- list(`Upregulated in Combo` = Combo_up,
                   `Downregulated in Combo` = Combo_down)
## Obtaing the raw counts from the OPC object
counts<- GetAssayData(seuObject_slim_nodoub, slot = "counts")
### celrankings 
cell_rankings<- AUCell_buildRankings(counts)
## Calculate AUCell for each cell
cells_AUC<- AUCell_calcAUC(AUCell_list, cell_rankings)
### cell assignmant
pdf("edgeR_new_exploration/deg/Unique_combo/AUCell/Thresholds.pdf")
cells_assignment<- AUCell_exploreThresholds(cells_AUC, plotHist = T, assign = T)
invisible(dev.off())
## Checking output
cells_assignment$geneSet$assignment

########### Adding the cells back in metadata 
auc_mtx <- t(getAUC(cells_AUC))          # cells x geneSets
colnames(auc_mtx) <- paste0(colnames(auc_mtx), "_AUC")
seuObject_slim_nodoub <- AddMetaData(seuObject_slim_nodoub, auc_mtx)
View(seuObject_slim_nodoub@meta.data) 
colnames(seuObject_slim_nodoub@meta.data)
## matirces of up and downregulation:: 0.1 and 0.075

########## Testing visualization #################
########################## #######################
feat <- "Downregulated in Combo_AUC"  # <- change as needed
thr  <- 0.075
col_bin <- paste0(feat, "_pos")

seuObject_slim_nodoub[[col_bin]] <- seuObject_slim_nodoub[[feat]] >= thr
#View(OPCs@meta.data)
p<-DimPlot(
  seuObject_slim_nodoub,
  group.by = col_bin,
  cols = c("FALSE" = "lightblue", "TRUE" = "#B2182B"), split.by = "Treatment"
) + guides(colour = guide_legend())
pdf("edgeR_new_exploration/deg/Unique_combo/AUCell/Downregulated_genes_Treatment.pdf", width = 10, height = 5)
p
dev.off()

```
```{r}
library(enrichplot)

emapplot(toppData, 
         showCategory = 30,
         layout = "kk")
```
####### GSEA and GO #####
################## ######
```{r}
library(readxl)
library(clusterProfiler)
library(org.Mm.eg.db)
Combo<- read_xlsx("edgeR_new_exploration/deg/Combo_DMSO/CTX_FullTab_shrink.xlsx")
#Combo_unique<- read_xlsx("edgeR_new_exploration/deg/Unique_combo/Combo_unique_CTX_DGE_shrink.xlsx")
Ribo<- read_xlsx("edgeR_new_exploration/deg/Ribo_DMSO/CTX_FullTab_shrink.xlsx")
BMS<- read_xlsx("edgeR_new_exploration/deg/BMS_DMSO/CTX_FullTab_shrink.xlsx")
Combo_Ribo<- read_xlsx("edgeR_new_exploration/deg/Combo_Ribo/CTX_FullTab_shrink.xlsx")
Combo_BMS<- read_xlsx("edgeR_new_exploration/deg/Combo_BMS/CTX_FullTab_shrink.xlsx")

### GSEA and GO one by one 
# Named numeric vector: logFC values with SYMBOL names
logFC <-Ribo$log2FoldChange
names(logFC) <- Ribo$Gene
# Map SYMBOL -> ENTREZID (mouse)
conv <- bitr(names(logFC),fromType = "SYMBOL",toType   = "ENTREZID",OrgDb = org.Mm.eg.db)
# Keep only mapped genes; collapse duplicates (ENTREZID repeated) by max abs(logFC)
df <- data.frame(SYMBOL = names(logFC), logFC = as.numeric(logFC))
df <- dplyr::inner_join(df, conv, by = "SYMBOL")
df <- df %>%dplyr::group_by(ENTREZID) %>%dplyr::summarise(logFC = logFC[which.max(abs(logFC))], .groups = "drop")
geneList <- df$logFC
names(geneList) <- df$ENTREZID
geneList <- sort(geneList, decreasing = TRUE)
#### Fixing gene list
geneList <- geneList[is.finite(geneList)]
geneList <- geneList[!is.na(names(geneList))]
geneList <- geneList[!duplicated(names(geneList))]
#geneList <- geneList[geneList != 0]
set.seed(708)
g <- gseGO(geneList = geneList,OrgDb= org.Mm.eg.db, keyType= "ENTREZID", ont= "BP", minGSSize=10, maxGSSize=500, pvalueCutoff = 1,verbose= FALSE, nPerm = 10000, eps= 0)
if (is.null(g) || nrow(g@result) == 0) return(NULL)
out <- g@result %>%dplyr::select(ID, Description, setSize, NES, pvalue, qvalue, core_enrichment) %>%dplyr::filter(qvalue < 0.05) #%>%dplyr::mutate(Cluster = paste0("Cluster ", ct))
out

#### Back into symbols 
ent2sym <- bitr(unique(unlist(strsplit(out$core_enrichment, "/"))),fromType = "ENTREZID",toType = "SYMBOL",OrgDb= org.Mm.eg.db)
sym_map <- setNames(ent2sym$SYMBOL, ent2sym$ENTREZID)
out2 <- out %>%mutate(core_symbols = vapply(core_enrichment, function(x) {ids <- strsplit(x, "/", fixed = TRUE)[[1]]
      syms <- unname(sym_map[ids])
      syms <- syms[!is.na(syms)]
      paste(unique(syms), collapse = "/")
    }, FUN.VALUE = character(1))
  )
out2
dir.create("edgeR_new_exploration/deg/Ribo_DMSO/GO_GSEA")
write_xlsx(out2, "edgeR_new_exploration/deg/Ribo_DMSO/GO_GSEA/BP_GO_GSEA.xlsx")
## viz 
top5_up <- out %>%filter(NES > 0) %>%slice_max(order_by = NES, n = 20, with_ties = FALSE) %>% ungroup()
top5_down <- out %>%filter(NES < 0) %>%slice_min(order_by = NES, n = 20, with_ties = FALSE) %>%ungroup()
#top5_both <- bind_rows(top5_up, top5_down) %>% mutate(Direction = ifelse(NES > 0, "Up", "Down"))
top5_both <- bind_rows(top5_up, top5_down) %>%mutate(Direction = ifelse(NES > 0, "Up", "Down"))
#top5_both <- top5_both %>%distinct(ID, .keep_all = TRUE)
top5_both <- bind_rows(top5_up, top5_down) %>%distinct(ID, .keep_all = TRUE) %>%arrange(NES) %>%                        mutate(Description = forcats::fct_inorder(Description))
###
pad <- 0.03 * diff(range(top5_both$NES, na.rm = TRUE))
p<-ggplot(top5_both, aes(x = NES, y = Description, fill = NES > 0)) + geom_col(width = 0.75) + geom_vline(xintercept = 0, linewidth = 0.4) +
  # GO IDs inside bars
  geom_text(aes(label = ID), hjust = ifelse(top5_both$NES > 0, 1.05, -0.05), color = "white", size = 4) +
  # Term names near x=0: left for positive, right for negative
  geom_text(aes(x = ifelse(NES > 0, -pad, pad),label = Description),hjust = ifelse(top5_both$NES > 0, 1, 0),size = 3) +
  tidytext::scale_y_reordered() +
  scale_fill_manual(values = c(`TRUE` = "firebrick", `FALSE` = "firebrick3")) + ##4575B4
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, size = 16),legend.position = "none",
    # remove y-axis line/ticks/labels since we now draw names ourselves
    axis.line.y  = element_blank(),axis.ticks.y = element_blank(),axis.text.y  = element_blank(),axis.text.x  = element_text(size = 14),axis.title.x = element_text(size = 16),
    # give room for the left/right text labels
    plot.margin = margin(5.5, 80, 5.5, 80)
  ) +
  coord_cartesian(clip = "off") +labs(x = "NES", y = NULL,title = "GSEA (BP) | Ribo vs DMSO")
pdf("edgeR_new_exploration/deg/Ribo_DMSO/GO_GSEA/BP_GO_GSEA_plot.pdf", width = 9, height = 7)
p
dev.off()
```
######### Progeny and decoupleR #####
#######################
```{r}
library(readxl)
library(progeny)
library(RColorBrewer)
library(decoupleR)
library(writexl)

counts(dds)
vsd <- vst(dds, blind = FALSE)
vst_mat <- assay(vsd)

### Normalized vst scores
vst_mat 
## Loading metadata
metadata<- read.csv("edgeR_new_exploration/metadata.csv")
metadata

### Downloading the get progeny from mice
net <- decoupleR::get_progeny(organism = 'mouse', top = 500)
#net1 <- progeny::getModel(organism = "Mouse", top = 500, decoupleR = TRUE)
net
write_xlsx(net, "edgeR_new_exploration/deg/decoupleR/get_progeny_mice.xlsx")

### Activity inference and progeny model
### Running mlm
sample_acts <- decoupleR::run_mlm(mat = vst_mat, net = net, 
                                  .source = 'source', 
                                  .target = 'target',
                                  .mor = 'weight', 
                                  minsize = 5)

sample_acts
## viz
sample_acts_mat <- sample_acts %>%
                   tidyr::pivot_wider(id_cols = 'condition', 
                                      names_from = 'source',
                                      values_from = 'score') %>%
                   tibble::column_to_rownames('condition') %>%
                   as.matrix()

# Scale per feature
sample_acts_mat <- scale(sample_acts_mat)

## viz
pal <- viridis::magma(200)

m <- max(abs(sample_acts_mat), na.rm = TRUE)
breaks <- seq(-m, m, length.out = length(pal) + 1)
pdf("edgeR_new_exploration/deg/decoupleR/decoupleR_Progeny_per_sample.pdf", height = 5, width = 10)
pheatmap::pheatmap(
  mat = t(sample_acts_mat),   
  fontsize = 14,
  fontsize_row = 12,
  fontsize_col = 12,
  color = pal,
  breaks = breaks,
  angle_col = 45,
  treeheight_col = 0,
  treeheight_row = 8,
  border_color = "grey60",
  cellwidth = 45,
  cellheight = 15,
  display_numbers = TRUE,
  fontsize_number = 13,
  number_color = "white"
)
dev.off()

### Progeny means per condition
###########################
sample_acts_mat
df_means <- as.data.frame(sample_acts_mat) %>%
  rownames_to_column("sample") %>%
  mutate(
    Condition = sub("_.*", "", sample),
    Condition = factor(Condition, levels = c("DMSO", "Ribo", "BMS", "Combo"))
  )
mean_per_condition <- df_means %>%
  dplyr::select(-sample) %>%
  group_by(Condition) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>%
  tibble::column_to_rownames("Condition")

## viz
pal <- viridis::magma(200)
m <- max(abs(mean_per_condition), na.rm = TRUE)
breaks <- seq(-m, m, length.out = length(pal) + 1)
pdf("edgeR_new_exploration/deg/decoupleR/decoupleR_Progeny_mean.pdf", height = 4, width = 6)
pheatmap::pheatmap(
  mat = t(mean_per_condition),
  cluster_cols = F,
  fontsize = 14,
  fontsize_row = 12,
  fontsize_col = 12,
  color = pal,
  breaks = breaks,
  angle_col = 45,
  treeheight_col = 0,
  treeheight_row = 8,
  border_color = "grey60",
  cellwidth = 45,
  cellheight = 15,
  display_numbers = TRUE,
  fontsize_number = 13,
  number_color = "white"
)
dev.off()

######################## ############ ############ ##################
### Inference of pathway activities from stat values in Combo vs DMSO
######################## ############ ############ ##################
resultsNames(dds)
res <- results(dds, name = "Condition_Combo_vs_DMSO")

### loading unique genes 
library(readxl)
unique <- read_xlsx("edgeR_new_exploration/deg/Unique_combo/Combo_unique_CTX_DGE_shrink.xlsx")

### Filtering res based to just get unique ones
unique_genes <- unique(unique$Gene)
unique_genes <- intersect(unique_genes, rownames(res))
res_unique <- res[unique_genes, ]
res_unique <- res_unique[order(res_unique$padj), ]
res_unique

## Adjust matrix
deg <- as.data.frame(res_unique) %>%
  rownames_to_column("ID") %>%
  dplyr::select(ID, stat) %>%
  filter(!is.na(stat)) %>%
  column_to_rownames("ID") %>%
  as.matrix()

contrast_acts <- decoupleR::run_mlm(mat = deg, 
                                    net = net, 
                                    .source = 'source', 
                                    .target = 'target',
                                    .mor = 'weight', 
                                    minsize = 5)
colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")[c(2, 10)])

p <- ggplot2::ggplot(data = contrast_acts, 
                     mapping = ggplot2::aes(x = stats::reorder(source, score), 
                                            y = score)) + 
     ggplot2::geom_bar(mapping = ggplot2::aes(fill = score),
                       color = "black",
                       stat = "identity") +
     ggplot2::scale_fill_gradient2(low = colors[1], 
                                   mid = "whitesmoke", 
                                   high = colors[2], 
                                   midpoint = 0) + 
     ggplot2::theme_minimal() +
     ggplot2::theme(axis.title = element_text(face = "bold", size = 12),
              axis.text.x = ggplot2::element_text(angle = 45, 
                                                  hjust = 1, 
                                                  size = 10, 
                                                  face = "bold"),
              axis.text.y = ggplot2::element_text(size = 10, 
                                                  face = "bold"),
              panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank()) +
     ggplot2::xlab("Pathways")

p
write_xlsx(contrast_acts, "edgeR_new_exploration/deg/decoupleR/Progeny/CombovsDMSO/UniquedecoupleR_progeny_CombovsDMSO.xlsx")

### Scatter plot visualization....
pathway <- c('TNFa')
df <- net %>%
      dplyr::filter(source == pathway) %>%
      dplyr::arrange(target) %>%
      dplyr::mutate(ID = target, 
                    color = "3") %>%
      tibble::column_to_rownames('target')
inter <- sort(dplyr::intersect(rownames(deg), rownames(df)))
df <- df[inter, ]
df['stat'] <- deg[inter, ]
df <- df %>%
      dplyr::mutate(color = dplyr::if_else(weight > 0 & stat > 0, '1', color)) %>%
      dplyr::mutate(color = dplyr::if_else(weight > 0 & stat < 0, '2', color)) %>%
      dplyr::mutate(color = dplyr::if_else(weight < 0 & stat > 0, '2', color)) %>%
      dplyr::mutate(color = dplyr::if_else(weight < 0 & stat < 0, '1', color))
colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")[c(2, 10)])
p1 <- ggplot2::ggplot(data = df, 
                     mapping = ggplot2::aes(x = weight, 
                                            y = stat, 
                                            color = color)) + 
     ggplot2::geom_point(size = 2.5, 
                         color = "black") + 
     ggplot2::geom_point(size = 1.5) +
     ggplot2::scale_colour_manual(values = c(colors[2], colors[1], "grey")) +
     ggrepel::geom_label_repel(mapping = ggplot2::aes(label = ID)) + 
     ggplot2::theme_minimal() +
     ggplot2::theme(legend.position = "none") +
     ggplot2::geom_vline(xintercept = 0, linetype = 'dotted') +
     ggplot2::geom_hline(yintercept = 0, linetype = 'dotted') +
     ggplot2::ggtitle(pathway)

p1

pdf("edgeR_new_exploration/deg/decoupleR/Progeny/CombovsDMSO/unique_decoupleR_barplot.pdf", height =5, width = 7)
p
dev.off()

pdf("edgeR_new_exploration/deg/decoupleR/Progeny/CombovsDMSO/unique_decoupleR_TGFb.pdf", height =5, width = 7)
p1
dev.off()
```
######## Trancription factor inference ########
################ ############### ##############
```{r}
## Transcription factors : also saved in Denis'file
net <- decoupleR::get_collectri(organism = 'mouse', 
                                split_complexes = FALSE)

net

# Run ulm
sample_acts <- decoupleR::run_ulm(mat = vst_mat, 
                                  net = net, 
                                  .source = 'source', 
                                  .target = 'target',
                                  .mor = 'mor', 
                                  minsize = 5)

n_tfs <- 25

# Transform to wide matrix
sample_acts_mat <- sample_acts %>%
                   tidyr::pivot_wider(id_cols = 'condition', 
                                      names_from = 'source',
                                      values_from = 'score') %>%
                   tibble::column_to_rownames('condition') %>%
                   as.matrix()

# Get top tfs with more variable means across clusters
tfs <- sample_acts %>%
       dplyr::group_by(source) %>%
       dplyr::summarise(std = stats::sd(score)) %>%
       dplyr::arrange(-abs(std)) %>%
       head(n_tfs) %>%
       dplyr::pull(source)

sample_acts_mat <- sample_acts_mat[,tfs]

# Scale per sample
sample_acts_mat <- scale(sample_acts_mat)

# Choose color palette
colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu"))
colors.use <- grDevices::colorRampPalette(colors = colors)(100)

my_breaks <- c(seq(-2, 0, length.out = ceiling(100 / 2) + 1),
               seq(0.05, 2, length.out = floor(100 / 2)))

# Plot
pheatmap::pheatmap(mat = sample_acts_mat,
                   color = colors.use,
                   border_color = "white",
                   breaks = my_breaks,
                   cellwidth = 15,
                   cellheight = 15,
                   treeheight_row = 20,
                   treeheight_col = 20)

#### Runnning contrats
########################
#########################
resultsNames(dds)
res <- results(dds, name = "Condition_Combo_vs_DMSO")
res

### loading unique genes 
library(readxl)
#unique <- read_xlsx("edgeR_new_exploration/deg/Unique_combo/Combo_unique_CTX_DGE_shrink.xlsx")

### Filtering res based to just get unique ones
unique_genes <- unique(unique$Gene)
unique_genes <- intersect(unique_genes, rownames(res))
res_unique <- res[unique_genes, ]
res_unique <- res_unique[order(res_unique$padj), ]
res_unique

## Adjust matrix
deg <- as.data.frame(res_unique) %>%
  rownames_to_column("ID") %>%
  dplyr::select(ID, stat, log2FoldChange, pvalue ) %>%
  filter(!is.na(stat)) %>%
  column_to_rownames("ID") %>%
  as.matrix()

str(deg)
contrast_acts <- decoupleR::run_ulm(mat = deg[, 'stat', drop = FALSE], 
                                    net = net, 
                                    .source = 'source', 
                                    .target = 'target',
                                    .mor='mor', 
                                    minsize = 5)

contrast_acts
contrast_acts<- contrast_acts |> filter( p_value < 0.5)

#### viz 
# Filter top TFs in both signs
f_contrast_acts <- contrast_acts %>%
                   dplyr::mutate(rnk = NA)

msk <- f_contrast_acts$score > 0

f_contrast_acts[msk, 'rnk'] <- rank(-f_contrast_acts[msk, 'score'])
f_contrast_acts[!msk, 'rnk'] <- rank(-abs(f_contrast_acts[!msk, 'score']))

tfs <- f_contrast_acts %>%
       dplyr::arrange(rnk) %>%
       head(n_tfs) %>%
       dplyr::pull(source)

f_contrast_acts <- f_contrast_acts %>%
                   filter(source %in% tfs)

colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")[c(2, 10)])

p <- ggplot2::ggplot(data = f_contrast_acts, 
                     mapping = ggplot2::aes(x = stats::reorder(source, score), 
                                            y = score)) + 
     ggplot2::geom_bar(mapping = ggplot2::aes(fill = score),
                       color = "black",
                       stat = "identity") +
     ggplot2::scale_fill_gradient2(low = colors[1], 
                                   mid = "whitesmoke", 
                                   high = colors[2], 
                                   midpoint = 0) + 
     ggplot2::theme_minimal() +
     ggplot2::theme(axis.title = element_text(face = "bold", size = 12),
              axis.text.x = ggplot2::element_text(angle = 45, 
                                                  hjust = 1, 
                                                  size = 10, 
                                                  face = "bold"),
              axis.text.y = ggplot2::element_text(size = 10, 
                                                  face = "bold"),
              panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank()) +
     ggplot2::xlab("TFs")

p

```
```{r}
str(deg)
tf <- 'Nr4a1'

df <- net %>%
      dplyr::filter(source == tf) %>%
      dplyr::arrange(target) %>%
      dplyr::mutate(ID = target, color = "3") %>%
      tibble::column_to_rownames('target')

inter <- sort(dplyr::intersect(rownames(deg), rownames(df)))

df <- df[inter, ]

df$logfc   <- deg[inter, "log2FoldChange"]
df$stat    <- deg[inter, "stat"]
df$p_value <- deg[inter, "pvalue"]

df <- df %>%
      dplyr::mutate(color = dplyr::if_else(mor > 0 & stat > 0, '1', color)) %>%
      dplyr::mutate(color = dplyr::if_else(mor > 0 & stat < 0, '2', color)) %>%
      dplyr::mutate(color = dplyr::if_else(mor < 0 & stat > 0, '2', color)) %>%
      dplyr::mutate(color = dplyr::if_else(mor < 0 & stat < 0, '1', color))

colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")[c(2, 10)])

p <- ggplot2::ggplot(data = df, 
                     mapping = ggplot2::aes(x = logfc, 
                                            y = -log10(p_value), 
                                            color = color,
                                            size = abs(mor))) + 
     ggplot2::geom_point(size = 2.5, 
                         color = "black") + 
     ggplot2::geom_point(size = 1.5) +
     ggplot2::scale_colour_manual(values = c(colors[2], colors[1], "grey")) +
     ggrepel::geom_label_repel(mapping = ggplot2::aes(label = ID,
                                                      size = 1)) + 
     ggplot2::theme_minimal() +
     ggplot2::theme(legend.position = "none") +
     ggplot2::geom_vline(xintercept = 0, linetype = 'dotted') +
     ggplot2::geom_hline(yintercept = 0, linetype = 'dotted') +
     ggplot2::ggtitle(tf)

p



```


######## Normal progeny scores #########
################### ########### ########
``` {r}
mt<- read_xlsx("edgeR_new_exploration/deg/gene_cpm.xlsx")
mt<- mt |> column_to_rownames("Gene")
mt<- as.matrix(mt)

progeny_df <- progeny(mt, scale=T, organism="Mouse", top=100, perm=1, 
    return_assay = TRUE)

pal <- magma(200)
m <- max(abs(progeny_mat), na.rm = TRUE)
breaks <- seq(-m, m, length.out = length(pal) + 1)

# samples as rows, pathways as columns
#pdf("edgeR_new_exploration/deg/Progeny_per_sample.pdf", height = 5, width = 10)
pheatmap(
  t(progeny_mat),
  fontsize = 14,
  fontsize_row = 12,
  fontsize_col = 12,
  color = pal,
  breaks = breaks,
  angle_col = 45,
  treeheight_col = 0,
  treeheight_row = 8,
  border_color = "grey60",
  cellwidth = 45,
  cellheight = 15,
  display_numbers = TRUE,
  fontsize_number = 13,
  number_color = "white"
)
#dev.off()
```

